Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN and qSVA normlaisation)

### 2017-08-16 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd
Added markdown to process CMC data with qSVA

---
title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId = "syn9872607",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = "syn9872607";
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';

thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'

@@

# Replace AFF and BP samples as other
levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")
```

@@

writeLines('Following samples are missing EV information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1)) 
```

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation
Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate

@@

```{r custom.normalisation, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate')

@@

### Differential expression analysis (Gender specific SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp1, fig.height=15, fig.width=20}
# Covariates to adjust
COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
postAdjustCovars = c('Dx.Tissue.Gender', 'Institution', 'RIN', 'IntronicRate')

@@

### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp2, fig.height=16, fig.width=8}
COVARIATES = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue')) %>%
  dplyr::mutate(Dx = factor(Dx, levels = c('SCZ', 'Control', 'Other')),
                Tissue = factor(Tissue))
postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')

@@

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN and qSVA normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd",
                                 parentId = "syn7501880",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - qSVA')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = "syn7501880";
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';

thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_qSVA.Rmd'

@@

# Replace AFF and BP samples as other
levels(METADATA$Dx) = c("Other", "Other", "Control", "SCZ")

# Get quality surrogate variables
qSV_DEG_MAT_DLPFC_ID = 'syn10284943'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = qSV_DEG_MAT_DLPFC_ID
qSV_DEG_MAT_DLPFC = downloadFile(qSV_DEG_MAT_DLPFC_ID)

qSV_DEG_MAT_ACC_ID = 'syn10284942'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = qSV_DEG_MAT_ACC_ID
qSV_DEG_MAT_ACC = downloadFile(qSV_DEG_MAT_ACC_ID)

qSV_DEG_MAT = rbindlist(list(qSV_DEG_MAT_DLPFC, qSV_DEG_MAT_ACC), use.names = T, fill = T)
tmp = t(qSV_DEG_MAT[,-(1)])
colnames(tmp) = qSV_DEG_MAT$V1
qSV_DEG_MAT = tmp[,colnames(tmp) %in% colnames(COUNT)]
rownames(qSV_DEG_MAT) = rownames(tmp)

qSV = sva::qsva(qSV_DEG_MAT)
colnames(qSV) = paste0('qSV', 1:dim(qSV)[2])

METADATA = METADATA %>%
  left_join(qSV %>% rownameToFirstColumn('SampleID'))
```

@@

writeLines('Following samples are missing EV information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1))

writeLines('Remove samples missing qSVs')
writeLines(paste('Following',length(METADATA$SampleID[is.na(METADATA$qSV1)]), 'samples are removed:'))
writeLines(paste(METADATA$SampleID[is.na(METADATA$qSV1)], collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!is.na(qSV1))
```

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5", "qSV1", "qSV2", "qSV3")

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>% 
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation
Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate, with qSV 1 to 3

@@

```{r custom.normalisation, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3')

@@

### Differential expression analysis (Gender specific SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp1, fig.height=15, fig.width=20}
# Covariates to adjust
COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
postAdjustCovars = c('Dx.Tissue.Gender', 'Institution', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3')

@@

### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp2, fig.height=16, fig.width=8}
COVARIATES = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue')) %>%
  dplyr::mutate(Dx = factor(Dx, levels = c('SCZ', 'Control', 'Other')),
                Tissue = factor(Tissue))
postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate', 'qSV1', 'qSV2', 'qSV3')

@@

  normalizationType	= 'qSVA'
)

# Code
CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - qSVA", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)
